Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-08 paper/main.tex
back to the short version
typos

%SHORT
% \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
% \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
% \captionsetup[figure]{format=myformat}
% \captionsetup[table]{format=myformat}
% \renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

%SHORT
% \lstset{basicstyle=\tt\scriptsize}
% \abovedisplayskip=3pt
% \belowdisplayskip=3pt
% \abovedisplayshortskip=3pt
% \belowdisplayshortskip=3pt
%SHORT

@@

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts: First, a parametric cost
  semantics formalizes the resource consumption of CompCert Clight
  programs.  Second, a quantitative Hoare logic enables users to
  interactively develop resource bounds in the Coq Proof Assistant.
  The logic is proved sound with respect to the cost semantics, and
  shallow embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for Clight computes derivations in the quantitative Hoare logic.
  It is the first automatic amortized analysis that can derive bounds
  that depend on negative integers or differences between numbers,
  which is crucial to handle typical systems code.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.
  %
  An experimental evaluation demonstrates the practicability of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging loop and recursion patterns
  from the literature and open-source software shows that the bounds
  derived by the automatic amortized analysis are often more precise.
\end{abstract}

@@

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver, and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publically available prototype implementation, and
  an experimental evaluation, that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Furthermore, the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during execution.

@@

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  Type-based amortized resource analysis works
well for functional programs with pattern matching but it has been a
long-time open problem to apply it to C-like programs with control
flow that depends integer arithmetic, negative numbers, and non-linear
control flow.
% example, it is a
% long-time open problem how to use type-based amortized analysis to
% derive bounds for functions whose resource consumption depends on
% (possibly negative) integers and non-sequential control flow as
% introduced by \code{break} and \code{return} statements.
% Unfortunately, these are the functions \iffull{with non-constant resource
% consumption}{} that are most common in CertiKOS and other system
% software.

@@

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program; a solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.

% To show the practicability of our approach we developed a case study
% with different real-time schedulers implemented in C.  To perform a
% schedulability analysis that takes into account the overhead of the
% scheduler we derived time bounds for context switching and
% initialization of the scheduler.  We then computed time bounds for a
% set of given user task and took account the previously computed
% scheduler overhead to determine whether the scheduling requirements of
% all user tasks can be met.

@@

To obtain a compositional reasnoning we also have to take into account
the state resulting from a program's execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for the resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

It is also possible to incorporate boolean preconditions into the
bounds to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre- and postconditions to take non-negative number or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  That
is, we use a resource metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

@@

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote a concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.22\linewidth}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y) 
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y = y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x = x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}

@@

$|[x, n]| + |[z, n]|$
\\[.4\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y = y+1;
  ${\color{blue}%
  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(3);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
${\color{blue}%
\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
while (y>9) {
  ${\color{blue}%
  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y-10;
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
   \end{lstlisting}

@@

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
that there are no restrictions on the signs of the input variables in
the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

@@

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

@@

As mentioned earlier, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
work~\cite{GulwaniMC09},  demonstrate that our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires heuristics for
its counter instrumentation.  It can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

@@

We include the built-in primitive $\code{assert}~E$ that terminates
the program if the argument $E$ evaluates to false and has no effect
otherwise.  This is useful to express assumptions on the inputs of a
program for the automatic analysis.  We also add the built-in function
$\code{tick}(n)$ that can be called with a constant integer $n$ as a
flexible way to model resource consumption or release (if $n$ is
negative).

>>>>

%SHORT
\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
\captionsetup[figure]{format=myformat}
\captionsetup[table]{format=myformat}
\renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

%SHORT
\lstset{basicstyle=\tt\scriptsize}
\abovedisplayskip=3pt
\belowdisplayskip=3pt
\abovedisplayshortskip=3pt
\belowdisplayshortskip=3pt
%SHORT

@@

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts: First, a parametric cost
  semantics formalizes the resource consumption of CompCert Clight
  programs.  Second, a quantitative Hoare logic enables users to
  interactively develop resource bounds in the Coq Proof Assistant.
  The logic is proved sound with respect to the cost semantics, and
  shallow embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for Clight computes derivations in the quantitative Hoare logic.
  It is the first automatic amortized analysis that can derive bounds
  that depend on negative integers or differences between numbers,
  which is crucial to handle typical systems code.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.
  %
  An experimental evaluation demonstrates the practicality of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging loop and recursion patterns
  from the literature and open-source software shows that the bounds
  derived by the automatic amortized analysis are often more precise.
\end{abstract}

@@

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver, and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publicly available prototype implementation, and
  an experimental evaluation, that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Furthermore, the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during execution.

@@

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  Type-based amortized resource analysis works
well for functional programs with pattern matching but it has been a
long-time open problem to apply it to C-like programs with control
flow that depends on integer arithmetic, negative numbers, and non-linear
control flow.
% example, it is a
% long-time open problem how to use type-based amortized analysis to
% derive bounds for functions whose resource consumption depends on
% (possibly negative) integers and non-sequential control flow as
% introduced by \code{break} and \code{return} statements.
% Unfortunately, these are the functions \iffull{with non-constant resource
% consumption}{} that are most common in CertiKOS and other system
% software.

@@

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects the resource cost and size changes of
integer variables in the program; a solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicality and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, and a linear
bound that amortizes the cost of $k$ successive increments to a binary
counter.

% To show the practicality of our approach we developed a case study
% with different real-time schedulers implemented in C.  To perform a
% schedulability analysis that takes into account the overhead of the
% scheduler we derived time bounds for context switching and
% initialization of the scheduler.  We then computed time bounds for a
% set of given user task and took account the previously computed
% scheduler overhead to determine whether the scheduling requirements of
% all user tasks can be met.

@@

To obtain a compositional reasnoning we also have to take into account
the state resulting from a program's execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for the resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

It is also possible to incorporate boolean conditions into a
bound to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre- and postconditions to take non-negative numbers or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  That
is, we use the \emph{tick metric} that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

@@

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.22\linewidth}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y) 
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y=y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x=x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}

@@

$|[x, n]| + |[z, n]|$
\\[.4\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y+1;
  ${\color{blue}%
  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(3);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
${\color{blue}%
\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
while (y>9) {
  ${\color{blue}%
  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y-10;
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
   \end{lstlisting}

@@

This idea is best explained by example.  If we again use the
tick metric that assigns cost $n$ to the function call
\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
that there are no restrictions on the signs of the input variables in
the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

@@

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in the rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $|x| + |y| + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-x-9$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

@@

As mentioned earlier, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
work~\cite{GulwaniMC09},  demonstrate that our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires heuristics for
its counter instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

@@

We include the built-in primitive $\code{assert}~e$ that terminates
the program if the argument $e$ evaluates to false and has no effect
otherwise.  This is useful to express assumptions on the inputs of a
program for the automatic analysis.  We also add the built-in function
\code{tick(n)} that can be called with a constant integer $n$ as a
flexible way to model resource consumption or release (if $n$ is
negative).
